---
title: "Alpha diversity"
author: "Mirna Vazquez Rosas Landa"
date: "2025-01-24"
output: html_document
---
## Data Import and Preprocessing

```{r setup, include=FALSE}
# Load necessary libraries
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(ggpubr)
library(gridExtra)
library(kableExtra)
library(tibble)
library(stringr)
library(pairwiseAdonis)
library(patchwork)
library(RColorBrewer)
library(writexl)

# Load data
load("../data/00_RData/ps.prevalence.green.RDATA")

# Verify that the data was loaded correctly
if (!exists("ps3")) {
  stop("The object 'ps3' is not loaded correctly. Check the file.")
}
```

# Alpha Diversity Analysis

We calculated multiple alpha diversity indices (Shannon, Simpson, and Chao1) and assessed their significance across factors such as season, zone, and depth.

This R script calculates and processes microbial community richness metrics from a phyloseq object (ps3) and prepares the data for further analysis:

## Richness Estimation:

The code calculates richness metrics (Observed, Chao1, Shannon, and Simpson) using the estimate_richness() function applied to the phyloseq object ps3.

## Error Handling:

If no richness metrics are generated (e.g., the result is an empty data frame), the code stops execution with an error message instructing the user to verify the ps3 object.

## Factor Conversion:

The variables depth, season, and zone from the metadata of the ps3 object are extracted and processed:

-depth is converted into a factor with specific levels (5, 20, 40).
-season is converted into a factor with levels (dry, flood).
-zone is extracted without modification.

## Data Cleaning:

Rows containing missing values (NA) in critical columns (zone, depth, season, and richness metrics) are removed using the filter() function from the dplyr package.

The resulting richness data frame is a clean, structured dataset that includes richness metrics and corresponding metadata (depth, season, and zone) for further analysis.

```{r richness}
# Estimate richness metrics (Observed, Chao1, Shannon, Simpson)
richness <- estimate_richness(ps3)

# Check if richness estimation was successful
if (ncol(richness) == 0) {
  stop("Richness metrics could not be estimated. Verify the 'ps3' object.")
}

# Convert 'depth', 'season', and 'zone' variables to factors
richness$depth <- factor(
  sample_data(ps3)$depth, levels = c("5", "20", "40")
)
richness$season <- factor(
  sample_data(ps3)$season, levels = c("dry", "flood")
)
richness$zone <- sample_data(ps3)$zone

# Remove rows with NA values in relevant columns
richness <- richness %>%
  filter(
    !is.na(zone), !is.na(depth), !is.na(season),
    !is.na(Observed), !is.na(Chao1), !is.na(Shannon), !is.na(Simpson)
  )
```

## Helper Functions

-perform_shapiro_tests

 -Purpose: Conduct Shapiro-Wilk tests for normality across multiple metrics and grouping variables.
 -Output: A data frame summarizing the p-values from Shapiro-Wilk tests.
 -Key Features: Dynamically pivots the dataset to a long format for testing. Tests are grouped by metric and specified grouping variables. Handles small sample sizes (n() < 3) by assigning NA.

-perform_kruskal_tests

 -Purpose: Perform Kruskal-Wallis tests to compare metrics across grouping factors (non-parametric).
 -Output: A summary of p-values for each metric and grouping factor.
 -Key Features:Constructs dynamic formulas like Value ~ zone for statistical testing.
Automatically generates columns for results using across() and .names.

-pairwise_comparisons

 -Purpose: Conduct pairwise Wilcoxon tests for a specific metric and grouping factor.
 -Output: A matrix of p-values adjusted using the Bonferroni method.

-extract_pvals

 -Purpose: Convert the p-values matrix from pairwise_comparisons into a clean and readable long format.
 -Output: A data frame summarizing pairwise comparisons, metrics, and adjusted p-values.
 -Key Features: Annotates significant p-values (p < 0.05) with an asterisk (*). Combines comparison pairs into a single column for clarity.

```{r setup, include=FALSE}
# Define helper functions
perform_shapiro_tests <- function(data, metrics, grouping_vars) {
  data %>%
    pivot_longer(cols = metrics, names_to = "Metric", values_to = "Value") %>%
    group_by(across(all_of(grouping_vars)), Metric) %>%
    summarise(Shapiro_p = ifelse(n() >= 3, shapiro.test(Value)$p.value, NA), .groups = "drop")
}

perform_kruskal_tests <- function(data, metrics, grouping_factors) {
  data %>%
    pivot_longer(cols = metrics, names_to = "Metric", values_to = "Value") %>%
    group_by(Metric) %>%
    summarise(
      across(
        all_of(grouping_factors),
        ~ {
          # Dynamically create a formula for kruskal.test
          grouping_var <- cur_column()
          formula <- as.formula(paste("Value ~", grouping_var))
          kruskal.test(formula, data = data)$p.value
        },
        .names = "Kruskal_p_{.col}"
      ),
      .groups = "drop"
    )
}

pairwise_comparisons <- function(data, metric, factor) {
  pairwise.wilcox.test(
    data[[metric]], data[[factor]], p.adj = "bonf", exact = FALSE
  )$p.value
}

extract_pvals <- function(pvals_list, metric, factor_name) {
  pvals <- as.data.frame(pvals_list)
  pvals <- pvals %>%
    rownames_to_column(var = "Comparison1") %>%
    pivot_longer(-Comparison1, names_to = "Comparison2", values_to = "p_value") %>%
    filter(!is.na(p_value)) %>%
    mutate(
      Factor = factor_name,
      Metric = metric,
      Comparison = paste(Comparison1, "vs.", Comparison2),
      p_value = ifelse(p_value < 0.05, paste0(formatC(p_value, digits = 3, format = "f"), "*"),
                       formatC(p_value, digits = 3, format = "f"))
    ) %>%
    select(Factor, Comparison, Metric, p_value)
  return(pvals)
}
```

## Data Cleaning and Preparation

Purpose: Remove rows with missing values in grouping variables or metrics to ensure clean data for analysis.
Output: A cleaned version of the dataset, richness_clean.

```{r data-cleaning}
# Define constants
metrics <- c("Observed", "Chao1", "Shannon", "Simpson")
grouping_vars <- c("zone", "depth", "season")

# Drop rows with NA values
richness_clean <- richness %>%
  drop_na(zone, depth, season, all_of(metrics))
```

## Shapiro-Wilk Normality Tests

  -Purpose: Test the normality of each metric grouped by the specified factors (zone, depth, season).
  -Output: A table showing Shapiro-Wilk test results.
  -Table Display: Rendered using kable() and kableExtra for a professional look.

```{r shapiro-tests}
# Perform Shapiro-Wilk normality tests
shapiro_results <- perform_shapiro_tests(richness_clean, metrics, grouping_vars)

# Display results
kable(shapiro_results, caption = "Shapiro-Wilk Normality Test Results", align = "c") %>%
  kable_styling(full_width = FALSE, position = "center")
```

## Kruskal-Wallis Tests

  -Purpose: Compare metrics across grouping factors using the Kruskal-Wallis test.
  -Output: A table summarizing the p-values for each metric and factor.
  -Table Display: Styled using kable() for clarity.

```{r kruskal-tests}
# Perform Kruskal-Wallis tests
kruskal_results <- perform_kruskal_tests(richness_clean, metrics, grouping_vars)

# Display results
kable(kruskal_results, caption = "Kruskal-Wallis Test Results", align = "c") %>%
  kable_styling(full_width = FALSE, position = "center")
```

## Pairwise Wilcoxon Tests

  -Purpose: Perform pairwise Wilcoxon tests to identify significant differences between groups for each metric.
  -Adjust p-values for multiple comparisons using the Bonferroni method.
  -Results Summary: The extract_pvals function formats the results into a human-readable summary. Pivoted to a wide format for a concise display.
  -Table Display:Styled using kable() and kableExtra to highlight significant results and include a footnote.

```{r pairwise-tests}
# Perform pairwise Wilcoxon tests
pairwise_pvals <- lapply(metrics, function(metric) {
  list(
    Zone = pairwise_comparisons(richness_clean, metric, "zone"),
    Depth = pairwise_comparisons(richness_clean, metric, "depth"),
    Season = pairwise_comparisons(richness_clean, metric, "season")
  )
})
names(pairwise_pvals) <- metrics

# Extract and format p-values
wilcox_results_summary <- bind_rows(
  lapply(metrics, function(metric) {
    bind_rows(
      extract_pvals(pairwise_pvals[[metric]]$Zone, metric, "Zone"),
      extract_pvals(pairwise_pvals[[metric]]$Depth, metric, "Depth"),
      extract_pvals(pairwise_pvals[[metric]]$Season, metric, "Season")
    )
  })
)

# Pivot table to wide format
wilcox_results_summary_wide <- wilcox_results_summary %>%
  pivot_wider(names_from = Metric, values_from = p_value) %>%
  arrange(Factor)

# Display the pairwise comparison results
kbl(wilcox_results_summary_wide,
    caption = "Pairwise Comparison p-values with Bonferroni Correction",
    align = "c", format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  add_footnote(label = "* Significant p-value (p < 0.05)", notation = "symbol")
```

## Supplementary tables
```{r}
# Create the Excel file with all tables
write_xlsx(
  list(
    "Richness Clean Data" = richness_clean,
    "Shapiro-Wilk Test Results" = shapiro_results,
    "Kruskal-Wallis Test Results" = kruskal_results,
    "Pairwise Comparison p-values" = wilcox_results_summary_wide
  ),
  path = "Analysis_Results.xlsx"
)
```

# Combined figure

- Reordering Factor Levels for 'zone': richness$zone <- factor(richness$zone, levels = c("Fringe", "Basin", "Impaired")). This reorders the factor levels of the zone variable to control the display order in subsequent plots.

-Creating Boxplots: For each diversity index (Observed, Shannon, Simpson, Chao1), the code creates a boxplot with ggplot2.
The boxplots show microbial diversity across zones (zone), with colors representing seasons (season). The plots also facet by depth (depth).
The geom_boxplot() function is used to create the boxplots, and facet_wrap(~depth) creates separate panels for each depth.
scale_fill_manual(values = c("darkorange","#D6ED17FF")) sets the color palette for the seasons.

-Wilcoxon Test Results: wilcox_results_summary contains results from pairwise Wilcoxon tests. p_value is cleaned and converted to numeric, and a new column (significance) is created, labeling results as "Significant" or "Not Significant" based on a threshold (0.05).

-Heatmap of Wilcoxon Test Results (p-values): A heatmap (p_value_heatmap) is generated using ggplot2 to visualize the significance of pairwise comparisons. The heatmap uses geom_tile() to fill cells with colors based on significance ("Significant" = green, "Not Significant" = gray).
The p-values are added as text labels with geom_text().
The heatmap is customized with axis labels, rotated text, and a legend indicating significance.

-Data Transformation (Pivoting and Reshaping): The combined_table is transformed from wide to long format using pivot_longer(), so each p-value column is converted into a row with its corresponding metric.
The significance column is redefined in the reshaped data based on the threshold, and Metric and Factor_Comparison are reordered for display in the heatmap.

-Second Heatmap (p_value_heatmap_2): A second heatmap (p_value_heatmap_2) is created using the reshaped data. It follows similar steps as the first heatmap but with slightly different settings, like adjusting text size for better readability.

-Combining All Plots: ggarrange() is used to combine the boxplots and heatmaps into a single figure. The ncol = 2, nrow = 3 argument arranges the plots in a 2x3 grid. Each plot is labeled with a letter (A-F) for easy reference. The common legend is placed at the bottom of the figure.

```{r alpha_diversity_visual}
# Reorder 'zone' factor levels to control the display order
richness$zone <- factor(richness$zone, levels = c("Fringe", "Basin", "Impaired"))

# 1. Boxplot for Shannon Index (Diversity), with Season and Zone
observed_plot <- ggplot(richness, aes(x = zone, y = Observed, fill = season)) +
  geom_boxplot() +
  facet_wrap(~depth) +
  theme_minimal() +
  labs(title = "Observed (Richness)", y = "Observed") +
  scale_fill_manual(values = c("darkorange","#D6ED17FF")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis for better readability

# 1. Boxplot for Shannon Index (Diversity), with Season and Zone
shannon_plot <- ggplot(richness, aes(x = zone, y = Shannon, fill = season)) +
  geom_boxplot() +
  facet_wrap(~depth) +
  theme_minimal() +
  labs(title = "Shannon Index (Diversity)", y = "Shannon Index") +
  scale_fill_manual(values = c("darkorange","#D6ED17FF")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis for better readability

# 2. Simpson Index plot (Evenness) with Season and Zone
simpson_plot <- ggplot(richness, aes(x = zone, y = Simpson, fill = season)) +
  geom_boxplot() +
  facet_wrap(~depth) +
  theme_minimal() +
  labs(title = "Simpson Index (Evenness)", y = "Simpson Index") +
  scale_fill_manual(values = c("darkorange","#D6ED17FF")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# 3. Chao1 Index plot (Richness) with Season and Zone
chao1_plot <- ggplot(richness, aes(x = zone, y = Chao1, fill = season)) +
  geom_boxplot() +
  facet_wrap(~depth) +
  theme_minimal() +
  labs(title = "Chao1 Index (Richness)", y = "Chao1 Index") +
  scale_fill_manual(values = c("darkorange","#D6ED17FF")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Define the significance threshold
threshold <- 0.05

# Convert p-value column to numeric and add Factor_Comparison
wilcox_results_summary <- wilcox_results_summary %>%
  mutate(p_value = as.numeric(gsub("[^0-9\\.]", "", p_value)),  # Clean and convert
         Factor_Comparison = paste(Factor, Comparison, sep = ":"),
         significance = ifelse(p_value < threshold, "Significant", "Not Significant"))

# Define order of metrics
ordered_metrics <- c("Observed", "Chao1", "Shannon", "Simpson")

# Reorder Metric factor levels
wilcox_results_summary$Metric <- factor(wilcox_results_summary$Metric, levels = ordered_metrics)

# Create the heatmap with binary color scale
p_value_heatmap <- ggplot(wilcox_results_summary, 
                          aes(x = Metric, y = Factor_Comparison)) +
  geom_tile(aes(fill = significance), color = "black") +  # Fill based on significance
  scale_fill_manual(values = c("Significant" = "green", "Not Significant" = "gray"), 
                    name = "Significance") +  # Manual binary color scale
  geom_text(aes(label = ifelse(!is.na(p_value), sprintf("%6.3f", p_value), "")), 
            size = 3, color = "black", hjust = 0.5, vjust = 0.5) +  # Add p-values as text
  theme_minimal() +
  labs(title = "Pairwise Wilcoxon Test Results (p-values)", 
       x = "Diversity Metric", y = "Comparison") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Rotate x-axis
        axis.text.y = element_text(size = 8),  # Adjust y-axis text size
        legend.position = "bottom") +  # Place legend at bottom
  guides(fill = guide_legend(override.aes = list(color = NA)))  # Clean up legend

# Print the heatmap
p_value_heatmap

# Convert to long format using pivot_longer from tidyr
long_format <- pivot_longer(combined_table, 
                            cols = starts_with("p_value_"), 
                            names_to = "Metric", 
                            values_to = "pvalue")


# Add a new column to classify significance
long_format_summary <- long_format %>%
  mutate(p_value = as.numeric(gsub("[^0-9\\.]", "", pvalue)),  # Clean and convert
         Factor_Comparison = paste(depth, Comparison, sep = ":"),
         significance = 
           ifelse(p_value < threshold, "Significant", 
                  "Not Significant")) %>%
  mutate(Metric=str_remove_all(Metric, "p_value_"))

ordered_comparison<- c("40 cm:Fringe vs Basin", "40 cm:Impaired vs Basin", "40 cm:Impaired vs Fringe",
                       "20 cm:Fringe vs Basin", "20 cm:Impaired vs Basin", "20 cm:Impaired vs Fringe",
                       "5 cm:Fringe vs Basin", "5 cm:Impaired vs Basin", "5 cm:Impaired vs Fringe")
# Reorder Metric factor levels
long_format_summary$Metric <- factor(long_format_summary$Metric, levels = ordered_metrics)
long_format_summary$Factor_Comparison <- factor(long_format_summary$Factor_Comparison, levels = ordered_comparison)


# Create the heatmap with binary color scale
p_value_heatmap_2 <- ggplot(long_format_summary, 
                          aes(x = Metric, y = Factor_Comparison)) +
  geom_tile(aes(fill = significance), color = "black") +  # Fill based on significance
  scale_fill_manual(values = c("Significant" = "green", "Not Significant" = "gray"), 
                    name = "Significance") +  # Manual binary color scale
  geom_text(aes(label = ifelse(!is.na(p_value), sprintf("%6.3f", p_value), "")), 
            size = 2, color = "black", hjust = 0.5, vjust = 0.5) +  # Add p-values as text
  theme_minimal() +
  labs(title = "Pairwise Wilcoxon Test Results (p-values)", 
       x = "Diversity Metric", y = "Comparison") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Rotate x-axis
        axis.text.y = element_text(size = 6),  # Adjust y-axis text size
        legend.position = "bottom") +  # Place legend at bottom
  guides(fill = guide_legend(override.aes = list(color = NA)))  # Clean up legend

# Print the heatmap
p_value_heatmap_2

# Combine all plots into one figure
combined_figure <- ggarrange(
  observed_plot, chao1_plot, shannon_plot, simpson_plot,  p_value_heatmap, p_value_heatmap_2,
  ncol = 2, nrow = 3,
  labels = c("A", "B", "C", "D", "E", "F"),
  common.legend = TRUE,
  legend = "bottom"
)

# Print the combined figure
print(combined_figure)

```
